# method {#chapter-label}  
:::frame
if you want to put your name to this, then feel free :) 
_Written by name_
:::

give a broad explanation of analysis. no more than a few paragraphs.
when should one use it?  
formulae?
assumptions?
etc.

## Packages used in this tutorial {-}

tidyverse or base R? e.g.

```{r message=FALSE,warning=FALSE}
library(tidyverse)
library(lme4)
```


## Research Question & Data {-}

explain the question, explain the data.
where does it come from? create a dictionary of the relevant variables:

```{r}
require(tidyverse)
tibble(
  variable_name = c("age","gender","height","pet"),
  description = c("age in years of the ppt",
                  "description...",
                  "description..",
                  "description...")
) %>% knitr::kable(.)
```

## Tidying {-}

if you want to include some specific data wrangling which you feel is an often needed skill for this sort of data, feel free to include it. 
alternatively, just import some tidy data and skip this bit 

## Checking assumptions {-}  

assumptions about data? 

## Analysis {-}

do step by step instructions of the analysis. 
try to include interpretations of each bit. 

## Checking assumptions {-}  

assumptions about model?

## Visualising {-}

show the code for making plots.

## Writing up {-}  

### methods section {-}

### results section {-}  

if there are nice tips/packages you use for getting results to look nice, include them.
e.g., memisc::mtable() for lme4 models to latex. 
don't worry about trying to cover all of word/latex/html etc. just do the one you know.  





