# R tips and tricks {#r-help}
```{r include=FALSE}
library(tidyverse)
```

### Trailing `%>%`s and `+`s {-}  

Having the last line of a piping sequence being a call to `print` has the benefit of making it possible to comment out lines without leaving a trailing pipe.  

This will leave `R` waiting: 
```{r eval=FALSE}
rnorm(10,10,1) %>% 
  round(2) %>% 
  #log() 
```

But this will not:
```{r}
rnorm(10,10,1) %>% 
  round(2) %>% 
  #log() %>%
  print
```

:::yellow
You can also use `print` wherever you want within a piping sequence, to simply print the output at that step to console (useful for identifying where piping sequences are going wrong!)  
:::

---

The same applies when using `ggplot`. Complete your ggplot with `+ NULL`, to ensure that you don't leave trailing `+`s when commenting out lines: 

```{r}
x=seq(-5,5,.1)
y=dnorm(x)

ggplot(mapping=aes(x=x,y=y))+
  geom_point()+
  geom_line()+
  theme_light()
```

This will leave R waiting for more...
```{r eval=FALSE}
ggplot(mapping=aes(x=x,y=y))+
  geom_point()+
  geom_line()+
  #theme_light()
```

But if we had finished our ggplot with `+NULL`, then commenting out that line will still render the plot!
```{r}
ggplot(mapping=aes(x=x,y=y))+
  geom_point()+
  geom_line()+
  #theme_light()+
  NULL
```



