---
title: "Lab 02"
author: "Dan Mirman"
output: html_document
---

# Basic MLR

## Exercise 1

Analyze the county-level suicide rate data from PHE (`PHE_MentalHealth.Rdata`) for 2001-2016

*  did the regions differ in their baseline (2001) suicide rates?
*  did the regions differ in ther slopes of change of suidice rate?
*  plot observed data and model fits

## Exercise 1 Solution

```{r}
library(tidyverse)
library(lme4)
library(effects)

load("../data/PHE_MentalHealth.Rdata") #load Public Health England data
unique(mh_phe[, 1:2]) #check list of mental health indicators: suicide is 41001

filter(mh_phe, IndicatorID == 41001) %>%
  ggplot(aes(Year, Value, color = Region)) +
  geom_smooth(method = "lm", se=FALSE)

# select data and shift Year variable so baseline year (2001) is 0
suicide_dat <- filter(mh_phe, IndicatorID == 41001) %>%
  mutate(Time = Year - 2001)
# base model: just change over time
m <- lmer(Value ~ Time + (Time | County), 
          data = suicide_dat, REML = F)
# add baseline differences between regions
m.0 <- lmer(Value ~ Time + Region + (Time | County), 
            data = suicide_dat, REML = F)
# add slope differences between regions
m.1 <- lmer(Value ~ Time * Region + (Time | County), 
            data = suicide_dat, REML = F)
# compare models
anova(m, m.0, m.1)
# regions differ in baseline suicide rate and in slope of change

# visualize effects
ef <- as.data.frame(effect("Time:Region", m.1))
ggplot(ef, aes(Time, fit, color=Region)) + 
  geom_line() +
  theme_bw() + scale_color_brewer(palette = "Set1")
```

## Exercise 2

Analyze the weight maintenance data (`WeightMaintain3`), a made-up data set based on Lowe et al. (2014, Obesity, 22, 94-100). Overweight participants completed a 12-week weight loss program, then were randomly assigned to one of three weight maintenance conditions:

* None (Control)
* MR (meal replacements): use MR to replace one meal and snack per day
* ED (energy density intervention): book and educational materials on purchasing and preparing foods lower in ED (reducing fat content and/or increasing water content of foods)

Weight was assessed at baseline (start of maintenance), 12 months post, 24 months post, and 36 months post.

(A) Overall, did the participants maintain their weight loss or gain weight?
(B) Did the groups differ in baseline weight change and rate of weight gain (non-maintenance)?
(C) Make a graph of the model fit
(D) Examine the parameter estimates and interpret them

## Exercise 2 Solution

```{r}
load("../data/WeightMaintain3.rda")
summary(WeightMaintain3)
```

(A) Overall, did the participants maintain their weight loss or gain weight?

```{r}
m.null <- lmer(WeightChange ~ 1 + (Assessment | ID), data=WeightMaintain3, REML=F)
m.base <- lmer(WeightChange ~ Assessment + (Assessment | ID), data=WeightMaintain3, REML=F)
anova(m.null, m.base)
```

Yes: $\chi^2(1)=56.5, p << 0.0001$

(B) Did the groups differ in baseline weight change and rate of weight gain (non-maintenance)?
```{r}
m.int <- lmer(WeightChange ~ Assessment + Condition + (Assessment | ID), data=WeightMaintain3, REML=F)
m.full <- lmer(WeightChange ~ Assessment*Condition + (Assessment | ID), data=WeightMaintain3, REML=F)
anova(m.null, m.base, m.int, m.full)
```

Yes: 

*  Baseline: $\chi^2(2)=9.4, p < 0.01$
*  Slope: $\chi^2(2)=40.4, p << 0.0001$

Note: `m.int` is difficult to interpret in light of massive effect on slope

```{r}
coef(summary(m.full))
```

Compared to no intervention, weight (re)gain was 1.75 lbs/year slower for the ED intervention and 0.84 lbs/year slower for the MR intervention.

Note that baseline weight difference parameters are not significantly different from 0.

(C) Make a graph of the model fit
```{r}
ggplot(WeightMaintain3, aes(Assessment, WeightChange, color=Condition)) + 
  stat_summary(fun.data=mean_se, geom="pointrange", size=1) + 
  stat_summary(aes(y=fitted(m.full)), fun.y=mean, geom="line") + 
  theme_bw(base_size=12) + scale_color_manual(values=c("black", "red", "blue"))

ggplot(fortify(m.full), aes(Assessment, WeightChange, color=Condition)) + 
  stat_summary(fun.data=mean_se, geom="pointrange", size=1) + 
  stat_summary(aes(y=.fitted), fun.y=mean, geom="line") + 
  theme_bw(base_size=12) + scale_color_manual(values=c("black", "red", "blue"))
```

(D) Examine the parameter estimates and interpret them

```{r}
round(coef(summary(m.full)), 3)
```

* `(Intercept)` ==> baseline weight change in None group
* `Assessment`  ==> slope of weight change in None group
* `ConditionED` ==> baseline weight change in ED group relative to None group
* `ConditionMR` ==> baseline weight change in MR group relative to None group
* `Assessment:ConditionED`  ==> slope of weight change in ED group relative to None group
* `Assessment:ConditionMR`  ==> slope of weight change in MR groups relative to None group

# Logistic MLR

## Exercise 3
In the `nwl` data set, the participants with aphasia are also separated into two groups based on the general location of their brain lesion: anterior vs. posterior. Compare these two groups:

(A) Is the learning rate (training data) different between these two groups?
(B) Does their test performance differ?
(C) Does their retention from immediate to follow-up test differ?

## Exercise 3 Solution
```{r}
load("../data/nwl.RData")
ggplot(filter(nwl, !is.na(lesion_location)), aes(block, PropCorrect, 
                                            color=lesion_location, 
                                            shape=lesion_location)) +
  stat_summary(fun.data=mean_se, geom="pointrange") + 
  stat_summary(data=filter(nwl, !is.na(lesion_location), block <= 7), 
                           fun.y=mean, geom="line") + 
  geom_hline(yintercept=0.5, linetype="dashed") + 
  geom_vline(xintercept=c(7.5, 8.5), linetype="dashed") + 
  scale_x_continuous(breaks=1:9, labels=c(1:7, "Test", "Follow-Up")) + 
  theme_bw(base_size=10) + 
  labs(x="Block", y="Proportion Correct", shape="Lesion\nLocation", color="Lesion\nLocation")
```

(A) Is the learning rate (training data) different between these two groups?
```{r}
m.base <- glmer(cbind(NumCorrect, NumError) ~ block + 
                  (block | ID), 
                data = filter(nwl, block < 8, !is.na(lesion_location)),
                family=binomial)
m.loc0 <- glmer(cbind(NumCorrect, NumError) ~ block + lesion_location + 
                  (block | ID), 
                data=filter(nwl, block < 8, !is.na(lesion_location)),
                family=binomial)
m.loc1 <- glmer(cbind(NumCorrect, NumError) ~ block * lesion_location + 
                  (block | ID), 
                data=filter(nwl, block < 8, !is.na(lesion_location)),
                family=binomial)
#summary(m.loc1)
anova(m.base, m.loc0, m.loc1)
```

(B) Does their test performance differ?
(C) Does their retention from immediate to follow-up test differ?

```{r}
m.recall.loc <- glmer(cbind(NumCorrect, NumError) ~ Phase * lesion_location + 
                        (Phase | ID), 
                  data=filter(nwl, block > 7, !is.na(lesion_location)),
                  family="binomial")
summary(m.recall.loc)
```