```{r, echo=FALSE}
HIDDEN_SOLS=FALSE
```

# Week 2: Basic Mixed Linear Regression  

`r msmbstyle::question_begin()`
Load the `tidyverse`, `lme4` and `effects` packages (install them if you haven't already).    
`r msmbstyle::question_end()` 
`r msmbstyle::solution_begin(hidden=HIDDEN_SOLS)`
```{r warning=FALSE,message=FALSE}
library(tidyverse)
library(lme4)
library(effects)
```
`r msmbstyle::solution_end()`

### Analyzing county-level suicide rate data from Public Health England (PHE).  

The data is available at []() and covers the period from 2001 to 2016. It contains information for a number of different indicators over this period for a selection of counties in England.  

`r msmbstyle::question_begin()`
  
* Did the regions differ in their baseline (2001) suicide rates?  
* Did the regions differ in ther slopes of change of suidice rate?  
* Make plots of the observed data and of model fits  
  
`r msmbstyle::question_end()`

`r msmbstyle::solution_begin(hidden=HIDDEN_SOLS)`
#### Loading, cleaning {-}  
```{r}
load("../labs/PHE_MentalHealth.Rdata") #load Public Health England data

unique(mh_phe[, 1:2]) #check list of mental health indicators: suicide is 41001

# plot observed data - regional rates
filter(mh_phe, IndicatorID == 41001) %>%
  ggplot(aes(Year, Value, color = Region)) +
  geom_smooth(method = "lm", se=FALSE)

# select data and shift Year variable so baseline year (2001) is 0
suicide_dat <- filter(mh_phe, IndicatorID == 41001) %>%
  mutate(Time = Year - 2001)
```

#### Modelling {-}
```{r}
# base model: just change over time
m <- lmer(Value ~ Time + (Time | County), 
          data = suicide_dat, REML = F)
# add baseline differences between regions
m.0 <- lmer(Value ~ Time + Region + (Time | County), 
            data = suicide_dat, REML = F)
# add slope differences between regions
m.1 <- lmer(Value ~ Time * Region + (Time | County), 
            data = suicide_dat, REML = F)

# compare models
anova(m, m.0, m.1)
```
  
It looks like regions differ in baseline suicide rate (addition of `Region` predictor in `m.0`) and in slope of change (addition of interaction in `m.1`)  

#### Plotting {-}  
```{r}
# you can extract the fitted values using the effect() function from the effects package.
ef <- as.data.frame(effect("Time:Region", m.1))
ggplot(ef, aes(Time, fit, color=Region)) + 
  geom_line() +
  theme_bw() + scale_color_brewer(palette = "Set1")
```
`r msmbstyle::solution_end()`




### Analyze the weight maintenance data (`WeightMaintain3`), a made-up data set based on Lowe et al. (2014, Obesity, 22, 94-100).  

Overweight participants completed a 12-week weight loss program, then were randomly assigned to one of three weight maintenance conditions:

* None (Control)  
* MR (meal replacements): use MR to replace one meal and snack per day  
* ED (energy density intervention): book and educational materials on purchasing and preparing foods lower in ED (reducing fat content and/or increasing water content of foods)  

Weight was assessed at baseline (start of maintenance), 12 months post, 24 months post, and 36 months post.  

```{r}
load("../labs/WeightMaintain3.rda")
summary(WeightMaintain3)
```

`r msmbstyle::question_begin()`
Overall, did the participants maintain their weight loss or gain weight?
`r msmbstyle::question_end()`
`r msmbstyle::solution_begin(hidden=HIDDEN_SOLS)`
```{r}
m.null <- lmer(WeightChange ~ 1 + (Assessment | ID), data=WeightMaintain3, REML=F)
m.base <- lmer(WeightChange ~ Assessment + (Assessment | ID), data=WeightMaintain3, REML=F)
anova(m.null, m.base)
```
Yes: $\chi^2(1)=56.5, p << 0.0001$
`r msmbstyle::solution_end()`

`r msmbstyle::question_begin()`
Did the groups differ in baseline weight change and rate of weight gain (non-maintenance)?
`r msmbstyle::question_end()`
`r msmbstyle::solution_begin(hidden=HIDDEN_SOLS)`
```{r}
m.int <- lmer(WeightChange ~ Assessment + Condition + (Assessment | ID), data=WeightMaintain3, REML=F)
m.full <- lmer(WeightChange ~ Assessment*Condition + (Assessment | ID), data=WeightMaintain3, REML=F)
anova(m.null, m.base, m.int, m.full)
```
Yes: 

*  Baseline: $\chi^2(2)=9.4, p < 0.01$
*  Slope: $\chi^2(2)=40.4, p << 0.0001$

Note: `m.int` is difficult to interpret in light of the massive effect on slope

```{r}
coef(summary(m.full))
```

Compared to no intervention, weight (re)gain was 1.75 lbs/year slower for the ED intervention and 0.84 lbs/year slower for the MR intervention.

Note that baseline weight difference parameters are not significantly different from 0.
`r msmbstyle::solution_end()`

`r msmbstyle::question_begin()`
Make a graph of the model fit
`r msmbstyle::question_end()`
`r msmbstyle::solution_begin(hidden=HIDDEN_SOLS)`
```{r}
ggplot(WeightMaintain3, aes(Assessment, WeightChange, color=Condition)) + 
  stat_summary(fun.data=mean_se, geom="pointrange", size=1) + 
  stat_summary(aes(y=fitted(m.full)), fun=mean, geom="line") + 
  theme_bw(base_size=12) + scale_color_manual(values=c("black", "red", "blue"))

ggplot(fortify(m.full), aes(Assessment, WeightChange, color=Condition)) + 
  stat_summary(fun.data=mean_se, geom="pointrange", size=1) + 
  stat_summary(aes(y=.fitted), fun=mean, geom="line") + 
  theme_bw(base_size=12) + scale_color_manual(values=c("black", "red", "blue"))
```
`r msmbstyle::solution_end()`


`r msmbstyle::question_begin()`
Examine the parameter estimates and interpret them
`r msmbstyle::question_end()`
`r msmbstyle::solution_begin(hidden=HIDDEN_SOLS)`
```{r}
round(coef(summary(m.full)), 3)
```

* `(Intercept)` ==> baseline weight change in None group
* `Assessment`  ==> slope of weight change in None group
* `ConditionED` ==> baseline weight change in ED group relative to None group
* `ConditionMR` ==> baseline weight change in MR group relative to None group
* `Assessment:ConditionED`  ==> slope of weight change in ED group relative to None group
* `Assessment:ConditionMR`  ==> slope of weight change in MR groups relative to None group
`r msmbstyle::solution_end()`


## Logistic MLR

## Effect of brain lesion location on learning rates    
```{r}
load("../labs/nwl.RData")
```

In the `nwl` data set, the participants with aphasia are also separated into two groups based on the general location of their brain lesion: anterior vs. posterior. Compare these two groups:

`r msmbstyle::question_begin()`
Is the learning rate (training data) different between these two groups?
**Tip:** Remember that you can use `cbind(nr_successes/nr_failures)` as an outcome in a binomial regression.  
`r msmbstyle::question_end()`
`r msmbstyle::solution_begin(hidden=HIDDEN_SOLS)`
```{r}
m.base <- glmer(cbind(NumCorrect, NumError) ~ block + (block | ID), 
                data = filter(nwl, block < 8, !is.na(lesion_location)),
                family=binomial)
m.loc0 <- glmer(cbind(NumCorrect, NumError) ~ block + lesion_location + (block | ID), 
                data=filter(nwl, block < 8, !is.na(lesion_location)),
                family=binomial)
m.loc1 <- glmer(cbind(NumCorrect, NumError) ~ block * lesion_location + (block | ID), 
                data=filter(nwl, block < 8, !is.na(lesion_location)),
                family=binomial)
#summary(m.loc1)
anova(m.base, m.loc0, m.loc1)
```
`r msmbstyle::solution_end()`

`r msmbstyle::question_begin()`
Does their test performance differ?
`r msmbstyle::question_end()`
`r msmbstyle::solution_begin(hidden=HIDDEN_SOLS)`

`r msmbstyle::solution_end()`

`r msmbstyle::question_begin()`
Does their retention from immediate to follow-up test differ?
`r msmbstyle::question_end()`
`r msmbstyle::solution_begin(hidden=HIDDEN_SOLS)`
```{r}
m.recall.loc <- glmer(cbind(NumCorrect, NumError) ~ Phase * lesion_location + 
                        (Phase | ID), 
                  data=filter(nwl, block > 7, !is.na(lesion_location)),
                  family="binomial")
summary(m.recall.loc)
```
`r msmbstyle::solution_end()`



`r msmbstyle::question_begin()`
**Extra:** Create a visualisation showing the differences between groups in the average proportion of correct responses at each point in time (i.e., each block, test, and follow-up)
`r msmbstyle::question_end()`
`r msmbstyle::solution_begin(hidden=HIDDEN_SOLS)`
```{r}
ggplot(filter(nwl, !is.na(lesion_location)), aes(block, PropCorrect, 
                                            color=lesion_location, 
                                            shape=lesion_location)) +
  stat_summary(fun.data=mean_se, geom="pointrange") + 
  stat_summary(data=filter(nwl, !is.na(lesion_location), block <= 7), 
                           fun=mean, geom="line") + 
  geom_hline(yintercept=0.5, linetype="dashed") + 
  geom_vline(xintercept=c(7.5, 8.5), linetype="dashed") + 
  scale_x_continuous(breaks=1:9, labels=c(1:7, "Test", "Follow-Up")) + 
  theme_bw(base_size=10) + 
  labs(x="Block", y="Proportion Correct", shape="Lesion\nLocation", color="Lesion\nLocation")
```
`r msmbstyle::solution_end()`
