```{r, echo=FALSE}
HIDDEN_SOLS=TRUE
```

# Other random effects structures

### Packages {-}  

```{r message=FALSE,warning=FALSE}
library(tidyverse)
library(lme4)
library(lmerTest)
library(effects)
```


### Lecture Slides {-}  

The lecture slides can be accessed [here](https://uoe-psychology.github.io/uoe_psystats/multivar/lectures/Lecture04_RanEf.html).  
Data from the lecture can be found at [https://edin.ac/36UYA00 (active_math_sim.csv)](https://edin.ac/36UYA00) and [https://edin.ac/2ShRCNl (problem_solving.Rdata)](https://edin.ac/2ShRCNl).  

### Background & Reading {-}  
  
+ [Baayen et al., 2008](https://doi.org/10.1016/j.jml.2007.12.005)  
+ [Barr et al., 2013](https://doi.org/10.1016/j.jml.2012.11.001)  
+ [Matuschek et al., 2017](https://doi.org/10.1016/j.jml.2017.01.001)  

  
## Crossed random effects  

### Exercise 1

<div class="red">
#### The data {-}  

```{r}
load(url("https://edin.ac/2ShRCNl"))
str(problem_solving)
```

* `Item`: word problem, can be *Hard* or *Easy*
* `Prob_Type`: difficulty level of word problem (16 hard problems, 14 easy problems)
* `Subject`: Participant ID, N=120
* `Condition`: whether the participant received the `Treatment` or not
* `RT`: time to solve the problem
* Note: there is some missing data because trials where the participants failed to solve the problem are excluded.

</div>

`r msmbstyle::question_begin(header = "&#x25BA; Question 1")`
Check the contrasts of the `Condition` and `Prob_Type` variables. 
`r msmbstyle::question_end()` 
`r msmbstyle::solution_begin(hidden=FALSE)`
```{r}
contrasts(problem_solving$Condition)
contrasts(problem_solving$Prob_Type)
```
They are sum-coded. Remember what this means when you are interpreting the intercept and coefficients!
`r msmbstyle::solution_end()`
  
<hr />

`r msmbstyle::question_begin(header = "&#x25BA; Question 2")`
+ Does `Prob_Type` vary within subjects?
+ Does `Prob_Type` vary within items?
+ Does `Condition` vary within subjects?
+ Does `Condition` vary within items?
`r msmbstyle::question_end()` 
`r msmbstyle::solution_begin(hidden=FALSE)`
+ Yes: Each subject solved both Easy and Hard problems.
+ No: Each item was either Easy or Hard.
+ No: Each subject was either in the control condition or the treatment condition.
+ Yes: Each item was solved both by subjects in the control condition and subjects in the treatment condition.
`r msmbstyle::solution_end()`
  
<hr />
  
`r msmbstyle::question_begin(header = "&#x25BA; Question 3")`
Conduct MLR with crossed random effects of subjects and items to answer these research questions

1. Did the treatment improve problem solving ability? 
2. Did the treatment effect differ between hard and easy problems?

**Hint:** You can do this with one model, given what you know about the contrasts.
`r msmbstyle::question_end()` 
`r msmbstyle::solution_begin(hidden=FALSE)`
```{r message=FALSE, warning=FALSE}
mod_ps <- lmer(RT ~ Prob_Type*Condition + 
                   (Prob_Type | Subject) + 
                   (Condition | Item), 
                 data=problem_solving, REML=FALSE)

summary(mod_ps)
```
+ Because both `Condition` and `Prob_Type` are sum-coded, the intercept term is the grand mean.  
+ The `Prob_Type1` coefficient corresponds to the mean for `Prob_Type == "Hard"` minus the grand mean.
<small>Note that sum-coding means that we no longer interpret this as the effect of `Prob_Type` at a reference level of `Condition` - instead it may help to think of it as the effect of `Prob_Type` 'when `Condition == 0`'. In the default coding, 0 corresponds to the reference level, but in sum-coding, where Control = 1 and Treatment = -1, what is 0? It's mid-way between the two (it may help to think of sum-coding similar to what happens when we 'center' continuous variables).</small>
+ The `Condition1` coefficient corresponds to the mean for `Condition == "Control"` minus the grand mean. 
+ The `Prob_Type1:Condition1` coefficient corresponds to the effect of `Prob_Type` when `Condition == "Control"` relative to the effect of `Prob_Type` at the grand mean. Or, equivalently, the effect of `Condition` when `Prob_Type == "Hard"` relative to the effect of `Condition` at the grand mean.  
<br>  
  
1. Treatment improves problem solving ability (slower reaction times in the control group across both easy and hard problems), $\beta = 58.34, SE = 29.43, p = 0.0499$
2. Treatment effect differs between easy and hard problems, with a greater effect for harder problems (greater difference in reaction times between control and treatment groups for hard problems), $\beta = 36.14, SE = 17.73, p = 0.0446$
  
`r msmbstyle::solution_end()`

<hr />

`r msmbstyle::question_begin(header = "&#x25BA; Question 4")`
Make a graph of the condition means and variability based on your analysis model.  
**Hint:** start with`as.data.frame(effect(........))`, and pass to it the parameter you're interested in and your model. 
`r msmbstyle::question_end()` 
`r msmbstyle::solution_begin(hidden=FALSE)`
```{r, fig.width=5, fig.height=4}
efx <- as.data.frame(effect("Prob_Type:Condition", mod_ps))

ggplot(efx, aes(Condition, fit, color=Prob_Type, ymin=fit-se, ymax=fit+se)) + 
  geom_pointrange() +
  theme_light() +
  scale_color_manual(values=c("lightgreen","tomato1")) +
  labs(y="Response Time", color="Problem\nType")
```
`r msmbstyle::solution_end()`

## Dealing with issue of non-convergence and singular fits  

<div class="lo">
### Singular fits {-}

You may have noticed that a lot of our models over the last few weeks have been giving a warning: `boundary (singular) fit: see ?isSingular`.   
Up to now, we've been largely ignoring these warnings. However, in this exercise we're going to look at how to deal with this issue.
<br>
The warning is telling us that our model has resulted in a 'singular fit'. Singular fits often indicate that the model is 'overfitted' - that is, the random effects structure which we have specified is too complex to be supported by the data.  

Perhaps the most intuitive advice would be remove the most complex part of the random effects structure (i.e. random slopes). This leads to a simpler model that is not over-fitted.  
Additionally, when variance estimates are very low for a specific random effect term, this indicates that the model is not estimating this parameter to differ much between the levels of your grouping variable. It might, in some experimental designs, be perfectly acceptable to remove this.  
<br>
A key point here is that when fitting a mixed model, we should think about how the data are generated. Asking yourself questions such as "do we have good reason to assume subjects might vary over time, or to assume that they will have different starting points (i.e., different intercepts)?" can help you in specifying your random effect structure
<br>
You can read in depth about what this means by reading the help documentation for `?isSingular`. For our purposes, a relevant section is copied below:  

*...intercept-only models, or 2-dimensional random effects such as intercept+slope models, singularity is relatively easy to detect because it leads to random-effect variance estimates of (nearly) zero, or estimates of correlations that are (almost) exactly -1 or 1.*

</div>



<div class="lo">
### Convergence warnings {-}

Issues of non-convergence can be caused by many things. If you're model doesn't converge, it does *not necessarily* mean the fit is incorrect, however it is **is cause for concern**, and should be addressed, else you may end up reporting inferences which do not hold.

<br>
There are lots of different things which you could do which *might* help your model to converge. A select few are detailed below:  

+ double-check the model specification and the data  
+ adjust stopping (convergence) tolerances for the nonlinear optimizer, using the optCtrl argument to [g]lmerControl. (see `?convergence` for convergence controls).  
+ center and scale continuous predictor variables (e.g. with `scale`)  
+ change the optimiser, or use `allFit()` to try the fit with all available optimizers. This will of course be slow, but is considered 'the gold standard'; *"if all optimizers converge to values that are practically equivalent, then we would consider the convergence warnings to be false positives."*  

</div>


### Exercise 2  

<div class = "red">
#### The data {-}  

An experiment was run to replicate "test-enhanced learning" (Roediger & Karpicke, 2006): two groups of 25 participants were presented with material to learn. One group studied the material twice (`StudyStudy`), the other group studied the material once then did a test (`StudyTest`). Recall was tested immediately (one minute) after the learning session and one week later. The recall tests were composed of 175 items identified by a keyword (`Test_word`).  
<br>  
The critical (replication) prediction is that the `StudyStudy` group should perform somewhat better on the immediate recall test, but the `StudyTest` group will retain the material better and thus perform better on the 1-week follow-up test.

```{r}
load(url("https://edin.ac/2RWbl6g"))
str(tel)
```
</div>

`r msmbstyle::question_begin(header = "&#x25BA; Question 5")`
Plot the data. Does it look like the effect was replicated?
`r msmbstyle::question_end()` 
`r msmbstyle::solution_begin(hidden=FALSE)`
You can make use of `stat_summary()` again!
```{r, eval=FALSE}
ggplot(tel, aes(Delay, Correct, col=Group)) + 
  stat_summary(fun.data=mean_se, geom="pointrange")+
  theme_light()
```
It's more work, but some people might rather calculate the numbers and then plot them directly. It does just the same thing: 
```{r}
tel %>% 
  group_by(Delay, Group) %>%
  summarise(
    mean = mean(Correct),
    se = sd(Correct)/sqrt(n())
  ) %>%
  ggplot(., aes(x=Delay, col = Group)) +
  geom_pointrange(aes(y=mean, ymin=mean-se, ymax=mean+se))+
  theme_light() +
  labs(y = "Correct")
```
  
That looks like test-enhanced learning to me!  
`r msmbstyle::solution_end()` 

<hr />

`r msmbstyle::question_begin(header = "&#x25BA; Question 6")`
Test the critical hypothesis using a mixed-effects model. Fit the maximal random effect structure supported by the experimental design.  

Some questions to consider:  
  
+ Item accuracy is a binary variable. What kind of model will you use?  
+ We can expect variability across subjects (some people are better at learning than others) and across items (some of the recall items are harder than others). How should this be represented in the random effects?
`r msmbstyle::question_end()` 
`r msmbstyle::solution_begin(hidden=FALSE)`
```{r}
m <- glmer(Correct ~ Delay*Group +
             (1 + Delay | Subject_ID) +
             (1 + Delay + Group | Test_word),
           data=tel, family="binomial",
           glmerControl(optimizer = "bobyqa"))
summary(m)
```
`r msmbstyle::solution_end()` 

`r msmbstyle::question_begin(header = "&#x25BA; Question 7")`
Your model with maximal random effects will probably not converge, or will obtain a singular fit. Simplify the model until you achieve convergence.  
<br>
What we're aiming to do here is to follow [Barr et al.'s](https://doi.org/10.1016/j.jml.2012.11.001) advice of defining our maximal model and then removing only the terms to allow a non-singular fit.  
<br>
**Note:** This strategy - starting with the maximal random effects structure and removing terms until obtaining model convergence, is just *one* approach, and there are drawbacks (see [Matuschek et al., 2017](https://doi.org/10.1016/j.jml.2017.01.001)). There is no consensus on what approach is best (see `?isSingular`).  
<br>
<br>
*Tip:* you can look at the variance estimates and correlations easily by using the `VarCorr()` function. What jumps out?  
<br>
*Hint:* Generalization over subjects could be considered more important than over items - if the estimated variance of slopes for `Delay` and `Group` by-items are comparatively small, it might be easier to remove them?  
`r msmbstyle::question_end()` 
`r msmbstyle::solution_begin(hidden=FALSE)`
```{r}
VarCorr(m)
```
The by-item slope of `Group` seems to be quite highly correlated with other by-item terms.   

For now, we will just simply remove the term (however, we could - if we had theoretical justification - constrain our model so that there was 0 correlation)
```{r}
m2 <- glmer(Correct ~ Delay*Group +
             (1 + Delay | Subject_ID) +
             (1 + Delay | Test_word),
            data=tel, family="binomial",
            glmerControl(optimizer = "bobyqa"))
VarCorr(m2)
```

It's still a singular fit, and the `Delay` random slope by `Test_word` variance is extremely low and perfectly correlated with the intercept, let's try removing that:
```{r}
m3 <- glmer(Correct ~ Delay*Group +
              (1 + Delay | Subject_ID) +
              (1 | Test_word),
            data=tel, family="binomial",
            glmerControl(optimizer = "bobyqa"))
```
Hooray, the model converged! 
```{r}
summary(m3)
```
`r msmbstyle::solution_end()` 

<!-- You can also get some odds-ratios and quick confidence intervals using the following: -->
<!-- *Note, the `method = "Wald"` corresponds to a quick-and-easy method of calculating confidence intervals (simply $1.96 \times SE$). The gold standard for confidence intervals is `method = "boot"`, which obtains them via bootstrapping, but this will take too long!  -->
<!-- ```{r} -->
<!-- exp(fixef(m4)) -->
<!-- exp(confint(m4, method="Wald"))[8:11,] -->
<!-- ``` -->
<!-- No difference between groups on the immediate test.  -->
<!-- For the `StudyStudy` group, there was a decreased likelihood of getting an item correct on the follow-up test relative to the immediate test ($OR = 0.35, \text{95% CI} = [0.31, 0.40]$).   -->
<!-- For the The likelihood of correctly answering an item  -->

<hr />

`r msmbstyle::question_begin(header = "&#x25BA; Question 8")`
Plot the model-estimated condition means and variability.  

*Hint:* This is very similar to question 4.
`r msmbstyle::question_end()` 
`r msmbstyle::solution_begin(hidden=FALSE)`
```{r}
ef <- as.data.frame(effect("Delay:Group", m3))

ggplot(ef, aes(Delay, fit, color=Group)) + 
  geom_pointrange(aes(ymax=upper, ymin=lower), position=position_dodge(width = 0.2))+
  theme_classic() # just for a change :)
```
`r msmbstyle::solution_end()` 

<hr />

`r msmbstyle::question_begin(header = "&#x25BA; Question 9")`
What should we do with this information? How can we apply test-enhanced learning to learning R and statistics?
`r msmbstyle::question_end()` 
`r msmbstyle::solution_begin(hidden=FALSE)`
You'll get the benefits of test-enhanced learning if you try yourself before looking at the solutions! If you don't test yourself, you're more likely to forget it in the long run. 
`r msmbstyle::solution_end()` 

### Exercise 3  

<div class="red">  
#### The data {-}  
  
Made-up data from a RCT treatment study: 5 therapists randomly assigned participants to control or treatment group and monitored the participants' performance over time. There was a baseline test, then 6 weeks of treatment, with test sessions every week (7 total sessions).

```{r}
load(url("https://edin.ac/2GPGgev"))
summary(tx)
```
</div> 

`r msmbstyle::question_begin(header = "&#x25BA; Question 10")`
Plot the data. Does it look like the treatment had an effect on the performance score?
`r msmbstyle::question_end()` 
`r msmbstyle::solution_begin(hidden=HIDDEN_SOLS)`
```{r message=FALSE}
ggplot(tx, aes(session, Score, color=group)) +
  stat_summary(fun.data = mean_se, geom="pointrange") +
  stat_smooth() +
  theme_classic()
```

Just for fun, let's add on the individual participant scores, and also make a plot for each therapist. 
```{r message=FALSE}
ggplot(tx, aes(session, Score, color=group)) +
  stat_summary(fun.data = mean_se, geom="pointrange") +
  stat_smooth() +
  theme_classic() +
  geom_line(aes(group=PID), alpha=.2)+facet_wrap(~therapist)
```
`r msmbstyle::solution_end()` 

<hr />

`r msmbstyle::question_begin(header = "&#x25BA; Question 11")`
Consider these questions when you're designing your model(s) and use your answers to motivate your model design and interpretation of results:  

a. What are the levels of nesting? How should that be reflected in the random effect structure?
a. What is the shape of change over time? Do you need polynomials to model this shape? If yes, what order polynomials?
`r msmbstyle::question_end()` 
`r msmbstyle::solution_begin(hidden=HIDDEN_SOLS)`
a. There are repeated measures of participants (session). There are also repeated measures of therapists (each one treated many participants).
a. Looks like linear change, don't need polynomials. Good to know that there is no difference at baseline, so no need for orthogonal time.
`r msmbstyle::solution_end()` 

<hr />

`r msmbstyle::question_begin(header = "&#x25BA; Question 12")`
Test whether the treatment had an effect using mixed-effects modeling. Specify the maximal model. 
`r msmbstyle::question_end()` 
`r msmbstyle::solution_begin(hidden=HIDDEN_SOLS)`
```{r}
# start with maximal model
m1 <- lmer(Score ~ session * group + 
             (1 + session | PID) + 
             (1 + session | therapist),
           data=tx, REML=FALSE)
summary(m1)
```
`r msmbstyle::solution_end()` 

<hr />

`r msmbstyle::question_begin(header = "&#x25BA; Question 13")`
Try adjusting your model by removing random effects or correlations, examine the model again, and so on..  
`r msmbstyle::question_end()` 
`r msmbstyle::solution_begin(hidden=HIDDEN_SOLS)`
```{r}
VarCorr(m1)
```
There's a correlation of exactly -1 between the random intercepts and slopes for therapists, and the standard deviation estimate for `session|therapist` is pretty small. Let's remove it. 
```{r}
m2 <- lmer(Score ~ session * group + 
             (1 + session | PID) + 
             (1 | therapist),
           data=tx, REML=FALSE)
VarCorr(m2)
```
It now looks like estimates for random intercepts for therapists is now 0. If we remove this, our model finally is non-singular:
```{r}
m3 <- lmer(Score ~ session * group + 
             (1 + session | PID),
           data=tx, REML=FALSE)
summary(m3)
```
Lastly, it's then a good idea to check that the parameter estimates and SE are not radically different across these models (they are virtually identical)
```{r}
summary(m1)$coefficients
summary(m2)$coefficients
summary(m3)$coefficients
```
`r msmbstyle::solution_end()` 


`r msmbstyle::question_begin(header = "&#x25BA; EXTRA: Question 14")`
Try the code below to use the `allFit()` function to fit your final model with all the available optimizers.  
  
+ You might need to install the `dfoptim` package to get one of the optimizers  
+ If you have an older version of `lme4`, then `allFit()` might not be directly available, and you will need to run the following: `source(system.file("utils", "allFit.R", package="lme4"))`. 

```{r eval=FALSE}
fits <- allFit(yourmodel)
summary(fits)
```
`r msmbstyle::question_end()` 
