```{r, echo=FALSE}
HIDDEN_SOLS=TRUE
set.seed(15732)
ggplot2::theme_set(ggplot2::theme_gray(base_size=13))

library(tidyverse)
```

# Paired T-Test {#chap-paired-mean-test}


<div class="lo">
#### Instructions {-}
  
- In this two-hour lab we will go through worked examples in the first hour, and you will attempt to answer some questions in the second hour.
- The Rmarkdown file for this week is [here](https://uoe-psychology.github.io/uoe_psystats/dapr1/labsheets/week_18_practice.Rmd).


#### Learning outcomes {-}

**LO1.** 

**LO2.** 

**LO3.** 

<!-- #### Reading {-} -->

</div>

## paired t-test  

### Paired data  

*pairs of time points*
Longitudinal Variables: Same person, same variable, different times

*pairs of variables*
Within-Person Contrast: Same person, different variables

*pairs of people*
Dyadic Data: Same variable, different people



### Example questions {-}



### Hypotheses {-}  


• The null hypothesis (H0) assumes that the true mean difference (μd) is equal to zero.
• The two-tailed alternative hypothesis (H1) assumes that μd is not equal to zero.
• The upper-tailed alternative hypothesis (H1) assumes that μd is greater than zero.
• The lower-tailed alternative hypothesis (H1) assumes that μd is less than zero.
The mathematical representations of the null and alternative hypotheses are defined below:

H0: μd = 0
H1: μd ≠ 0    (two-tailed)
H1: μd > 0    (upper-tailed)
H1: μd < 0    (lower-tailed)


**Null hypothesis:**  


**Alternative hypothesis:** 



### Formula {-}  

one sample t test 
$$
t = \frac{\bar{x} - \mu_0}{\frac{s}{\sqrt{n}}}
$$

two independent samples t test 
$$
t = \frac{\bar{x_1} - \bar{x_2}}{\sqrt{\frac{s_1^2}{n_1}+\frac{s_2^2}{n_2}}}
$$

paired samples t test 
$$
t = \frac{\bar{d} - 0}{\frac{s_d}{\sqrt{n}}}
$$

Where $d$ is the difference between pairs $d = x_1 - x_2$,  
$\bar{d}$ is the mean of the differences,  
and $s_d$ is the standard deviation of the differences.  

With the one sample t test, the steps we do are:

1. calculate the sample mean
1. calculate the sample standard deviation 
1. calculate t
1. calculate the probability of observing a t statistic at least as extreme assuming the null hypothesis to be true. 

With the paired sample t test, the steps we are:

**0. calculate the difference in pairs**
1. calculate the sample mean **difference**
1. calculate the sample standard deviation **of differences**
1. calculate t
1. calculate the probability of observing a t statistic at least as extreme assuming the null hypothesis to be true. 


and it's that easy! the remainder is the same as we did in [Week 16](#chap-one-mean-test).


## Walkthrough 


- scores on ACE-III 
The Addenbrooke's Cognitive Examination-III (ACE-III) is a brief cognitive test that assesses five cognitive domains: attention, memory, verbal fluency, language and visuospatial abilities. ... The total score is 100 with higher scores indicating better cognitive functioning

q - how is the data paired?
a - same person, same variable, different time points

q hypotheses?

q - make a new column of the difference in scores between time points  
a - mutate(diff = t2-t1)

q - calculate $\bar{x}_d$ and $s_d$.  
a - summarise(mean(diff), sd(diff))

q - fill in blanks t-statistic 

q - calculate t in r

q - calculate pr(t)  

q - how would we report this? t(df) = , p </>/= ?


## ALL IN ONE LINE OF CODE! {-}

q - run ?t.test. what might be the argument we need to use for this test?  
a - paired = TRUE  



## Assumptions {-}  

 The dependent variable must be continuous (interval/ratio).
• The observations are independent of one another.
• The dependent variable should be approximately normally distributed.
• The dependent variable should not contain any outliers.


## Summary  



## Lab  
  
_Please attempt the questions **before** looking at the solutions. Copy & pasting the solutions will not help with learning!_  



data - age differences in heterosexual marriages
[here](http://www.lock5stat.com/datasets/MarriageAges.txt)
```{r}
marriages<-read.table("http://www.lock5stat.com/datasets/MarriageAges.txt", header = TRUE)
```


q - add new column which is the husband's age minus the wife's age
```{r}
marriages <- marriages %>% 
  mutate(
    agediff = Husband - Wife
  )


```


q - think. if we get a positive t statistic, t>0, what direction is the difference? who tends to be older than who? 
a - because we did husband minus wife, a positive t statistic indicates men tend to be older than women in heterosexual marriages. 

q - following the procedure we learnt in week 16, calculate the t statistic manually. 

```{r}
terms <- marriages %>% 
  summarise(
    xbar = mean(agediff),
    s = sd(agediff),
    mu_0 = 0,
    n = n()
  )

t = (terms$xbar - terms$mu_0) / (terms$s / sqrt(terms$n))
t
```



data - school students preference for math vs preference for english  

data - scores pre and post intervention  

data - practice effects. 














### Glossary  
