```{r, echo=FALSE}
HIDDEN_SOLS=TRUE
TOGGLE=TRUE
set.seed(15732)
ggplot2::theme_set(ggplot2::theme_gray(base_size=13))

library(tidyverse)
```

# Chi-squar tests {#chap-chi-square}


<div class="lo">
#### Instructions {-}
  
- In this two-hour lab we will go through worked examples in the first hour, and you will attempt to answer some questions in the second hour.
- The Rmarkdown file for this week is [here](https://uoe-psychology.github.io/uoe_psystats/dapr1/labsheets/week_19_practice.Rmd).


#### Learning outcomes {-}

By the end of this lab, you should be able to:

**LO1.** Identify the appropriate test for the task at hand: association or homogeneity.

**LO2.** Check the validity conditions.

**LO3.** Report the test results.

</div>



## Recap {-}



## Walkthrough: Birthdays of the week


## Lab: 

```{r echo=FALSE}
library(tidyverse)

set.seed(15732)


sports <- tibble(watched = c(
    rep('Yes', 47), rep('No', 21),
    rep('Yes', 26), rep('No', 23),
    rep('Yes', 14), rep('No', 37)
  ),
  affected = c(
    rep('Primary', 47), rep('Primary', 21),
    rep('Somewhat', 26), rep('Somewhat', 23),
    rep('No', 14), rep('No', 37)
  ))

sports$watched <- factor(sports$watched, levels = c('Yes', 'No'))
sports$affected <- factor(sports$affected, levels = c('Primary', 'Somewhat', 'No'))

sports <- sports[sample.int(nrow(sports)), ]

table(sports)

write_csv(sports, 'sport_university.csv', col_names = TRUE)
```

The file `sport_university.csv` contains data on 168 university students. Each of them was asked whether they were watching sports when growing up (Yes/ No) and whether their university choice was influenced by the university's sports team.

```{r}
sports <- read_csv('sport_university.csv', col_names = TRUE)
sports

table(sports)
```

```{r}
mosaicplot(affected ~ watched, data = sports)
```


```{r}
sports_summaries <- sports %>%
  table %>%
  as.data.frame()
sports_summaries

ggplot(sports_summaries, aes(x = affected, y = Freq, fill = watched)) +
  geom_bar(stat = 'identity', position = 'fill')

ggplot(sports_summaries, aes(x = watched, y = Freq, fill = affected)) +
  geom_bar(stat = 'identity', position = 'fill')
```



## Coffee Drinking and Sexual Habits

The School of Divinity and the School of Psychology of University A and University B competed on a literary exam. 

```{r}

```



The data are as follows:



```{r}

```

