# Introduction {#intro}

<div class="lo">
#### Learning outcomes {-}

1. First
2. Second
3. Third
</div>


test questions.	
the code is in dropdowns.R file, which is sourced at the start of the bookdown.	
the .css and toggling script is included in toggle.html, which is then incuded in_header in the yaml

`r qbegin()`	
This is the first question.	
`r qend()`	

`r solbegin()`	
Hello, i'm a solution.	
`r solend()`	

`r qbegin()`	
Question again	
`r qend()`	

`r solbegin()`	
The solution is peppa pig. 	

I think these will just number through the whole book, so it won't reset each chapter, which is not ideal. 	
`r solend()`	

:::green	
This is a green box	
:::	

:::frame	
This is a frame :)	
:::	


You can label chapter and section titles using `{#label}` after them, e.g., we can reference Chapter \@ref(intro). If you do not manually label them, there will be automatic labels anyway, e.g., Chapter \@ref(methods).

Figures and tables with captions will be placed in `figure` and `table` environments, respectively.

```{r nice-fig, fig.cap='Here is a nice figure!', out.width='80%', fig.asp=.75, fig.align='center'}
par(mar = c(4, 4, .1, .1))
plot(pressure, type = 'b', pch = 19)
```

Reference a figure by its code chunk label with the `fig:` prefix, e.g., see Figure \@ref(fig:nice-fig). Similarly, you can reference tables generated from `knitr::kable()`, e.g., see Table \@ref(tab:nice-tab).

```{r nice-tab, tidy=FALSE}
knitr::kable(
  head(iris, 20), caption = 'Here is a nice table!',
  booktabs = TRUE
)
```

You can write citations, too. For example, we are using the **bookdown** package [@R-bookdown] in this sample book, which was built on top of R Markdown and **knitr** [@xie2015].
